Implications of afforestation policies for global markets

### 2024-02-20 paper.Rmd
Update paper.Rmd

---
title: "Implications of afforestation policies for global markets"
date: "`r format(Sys.time(), '%B %d, %Y')`"
author: "Eric Davis, Maros Ivanic, Brent Sohngen"
output: 
  bookdown::pdf_document2:
    keep_tex: TRUE
toc: FALSE
bibliography: references.bib
header-includes:
    - \usepackage{setspace}\doublespacing
abstract: "The projected growth of the population and incomes as implied by various shared socioeconomic pathways is expected to create  pressures on converting forests into farmland as food prices increase. Depending on the specific path taken, SSP1 through SSP5, and the continuation of the current trends in agricultural productivity growth, the market impacts are expected to vary but generally result in increases in food prices. Policies aimed at reversing the deforestation trend by afforesting cropland would result in severe negative market impacts such as  higher food prices, especially in low agricultural productivity/high population growth scenarios."
---

@@

The standard GTAP model represents a leading CGE model of the global economy with a specific focus on trade linkages. Its general equilibrium nature means that all markets that are considered in the model are cleared with market-clearing prices. The extensive coverage of sectors---the latest version of the GTAP database includes 68 sectors---makes it difficult to model sectors individually; instead the model applies nested CES production functions to all sectors, differentiating them by parameters only. To model one sector, e.g., forestry, differently from the rest would require major redesign of the model. While specific versions of the GTAP model are definitely possible but extremely common as they require to be be fully parametrized for all countries, which makes the option of adapting the GTAP model for a one-off application impractical.  To analyze carbon sequestration in GTAP, which would require a lot of effort, we instead create a connection between the standard GTAP model and the GTM model which has been specifically designed to analyze the relationship among land rents, forested land cover and carbon sequestration through forests, among other things.

Establishing a connection between the standard GTAP model and the GTM model is a two-step process. First, we decide on the set of common variables between the two models in those markets that are included in both models and their closure, i.e., which variables will be determined endogenously by one model and held as exogenous for the other model. In our case, we selected the return on land to be modeled by the GTAP model and fed into the GTM model as exogenous, while the amount of land available for crop and livestock production, while exogenous to the GTAP model, is endogenously determined by the GTM model. In the second, step we establish the connection between the models on the software level. This step requires that the GTAP model, written in GEMPACK can operate with the inputs from the GTM model, written in GAMS, and submit the outputs back to the GTM model. In order to allow for the two quite dated technologies, GEMPACK and GAMS, which were developed long time age before software integration become important or possible, we created a set of user functions in R that allow executing both models, collecting their outputs and an creating  parameter files for both systems automatically.

>>>>

---
title: "Implications of afforestation policies for global markets"
date: "`r format(Sys.time(), '%B %d, %Y')`"
author: "Eric C. Davis, Maros Ivanic, and Brent Sohngen"
output: 
  bookdown::pdf_document2:
    keep_tex: TRUE
toc: FALSE
bibliography: references.bib
header-includes:
    - \usepackage{setspace}\doublespacing
abstract: "The projected growth of the population and incomes as implied by various shared socioeconomic pathways is expected to create  pressures on converting forests into farmland as food prices increase. Depending on the specific path taken, SSP1 through SSP5, and the continuation of the current trends in agricultural productivity growth, the market impacts are expected to vary but generally result in increases in food prices. Policies aimed at reversing the deforestation trend by afforesting cropland would result in severe negative market impacts such as  higher food prices, especially in low agricultural productivity/high population growth scenarios."
---

@@

The standard GTAP model is a leading CGE representation of the global economy and its trade linkages. Its general equilibrium nature means that all markets that are considered in the model are cleared with market-clearing prices. There are 68 market sectors covered in the latest version of the GTAP database includes  sectors. As the model applies nested CES production functions to all sectors, differentiating them by parameters only, it is difficult to model a sector individually. To model one sector, in this case forestry, differently from the rest would require a major redesign of the model. While not impossible, we chose to create a connection between the standard GTAP model and the GTM model, as GTM has been specifically designed to analyze the relationship among land rents, forested land cover, and carbon sequestration through forests. Doing so obviated the need to reparametrize the GTAP model for all countries.

Establishing a connection between the standard GTAP model and the GTM model was a two-step process. First, we identified the set of common variables that are included in both models and their closures. That is, we selected which variables would be determined endogenously by one model and held as exogenous for the other model. In our case, we selected the return on land to be modeled by the GTAP model and fed into the GTM model as exogenous, while the amount of land available for crop and livestock production, while exogenous to the GTAP model, is endogenously determined by the GTM model. In the second, step we establish the connection between the models on the software level. This step requires that the GTAP model, written in GEMPACK can operate with the inputs from the GTM model, written in GAMS, and submit the outputs back to the GTM model. In order to allow for the two quite dated technologies, GEMPACK and GAMS, which were developed long time age before software integration become important or possible, we created a set of user functions in R that allow executing both models, collecting their outputs and an creating  parameter files for both systems automatically.
